---
title: "NYPD Shooting Incident Data Report"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE
                      , warning = FALSE
                      , message = FALSE
                      , fig.width = 12
                      , fig.height = 7)
```

## Table of Content

1.  [Introduction](#chapter_1)
2.  [Exploratory data analysis](#chapter_2)
    -   2.1 [Probability of shooting incident by borough](#chapter_2_1)
    -   2.2 [Circular heatmap of shooting incidents by hour
        range](#chapter_2_2)
3.  [Model](#chapter_3)
    -   3.1 [Summary](#chapter_3_1)
    -   3.2 [Performance](#chapter_3_2)
4.  [Bias](#chapter_4)
    -   4.1 [Possible bias](#chapter_4_1)
    -   4.2 [Dataset bias](#chapter_4_1)
5.  [Conclusion](#chapter_5)

## 1. Introduction <a class="anchor" id="chapter_1"></a>

The NYPD Shooting Incident Data (Historic) dataset includes records of
shooting incidents in NYC from 2006 to 2023.\
The dataset consists of **28,562** records with the following
attributes:

-   **INCIDENT_KEY:**\
    A unique identifier for each shooting incident.

-   **OCCUR_DATE:**\
    Details about date the incident took place.

-   **OCCUR_TIME:**\
    Details about time the incident took place.

-   **BORO:**\
    The borough of New York City where the incident occurred.

-   **LOC_OF_OCCUR_DESC:**\
    Categorizes the type of location (e.g., inside, outside).

-   **PRECINCT:**\
    The police precinct responsible for the area where the incident
    occurred.

-   **JURISDICTION_CODE:**\
    Indicates the jurisdiction under which the incident falls.

-   **LOC_CLASSFCTN_DESC:**\
    Categorizes the type of location (e.g., residential, commercial).

-   **LOCATION_DESC:**\
    Categorizes the type of location (e.g., supermarket, bank).

-   **STATISTICAL_MURDER_FLAG:**\
    A boolean flag indicating whether the incident resulted in murder.

-   **Demographic Information:**\
    Includes age group, sex, and race/ethnicity of both perpetrators and
    victims.

-   **Geographical Coordinates:**\
    Includes the latitude and longitude of the incident.

```{r}
library("tidyverse")
library("ggplot2")
library("fmsb")
library("httr")
library("readr")
```

```{r}
DATA_URL = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
response <- GET(DATA_URL, config = config(ssl_verifypeer = FALSE))

data <- read_csv(content(response, "text"), show_col_types = FALSE)
```

```{r}
summary(data)
```

## 2. Exploratory data analysis <a class="anchor" id="chapter_2"></a>

### 2.1 Probability of shooting incident by borough <a class="anchor" id="chapter_2_1"></a>

```{r}
data_boro = data %>%
    group_by(BORO) %>%
    summarise(INCIDENT_COUNT = n()) %>%
    mutate(INCIDENT_PROBABILITY = round(INCIDENT_COUNT/sum(INCIDENT_COUNT)*100,2)) %>%
    arrange(desc(INCIDENT_PROBABILITY))

data_boro
```

```{r}
ggplot(data_boro, aes(x = reorder(BORO, INCIDENT_PROBABILITY)
                      , y = INCIDENT_PROBABILITY)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = sprintf("%.2f", INCIDENT_PROBABILITY)), 
            hjust = -0.5, size = 3.5) +
            labs(title = "Probability of Shooting Incident by Borough",
                    x = "Borough",
                    y = "Probability (%)",
                    subtitle = paste("Total:"
                             , sum(data_boro$INCIDENT_COUNT))
            ) +
            theme_minimal() +
            ylim(0, max(data_boro$INCIDENT_PROBABILITY) * 1.1) +
            coord_flip()
```

### 2.2 Circular heatmap of shooting incidents by hour range <a class="anchor" id="chapter_2_2"></a>

```{r}
data_hour <- data %>%
  mutate(hour = cut(as.numeric(OCCUR_TIME)
                     , breaks = seq(0, 86400, by = 3600)
                     , labels = paste0(sprintf("%02d:00", seq(0, 23))
                                         , " - "
                                         , sprintf("%02d:00", seq(1, 24))
                     )
                     , right = FALSE
                   )
        ) %>%
  group_by(hour) %>%
  summarise(INCIDENT_COUNT = n(), .groups = 'drop')
```

```{r}
circular_heatmap <- ggplot(data_hour, aes(x = hour, y = "", fill = INCIDENT_COUNT)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "blue") +
    labs(title = "Circular Heatmap of Shooting Incidents by Hour Range",
         x = NULL,
         y = NULL) +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    coord_polar(theta = "x")
```

```{r}
print(circular_heatmap)
```

## 3. Model <a class="anchor" id="chapter_3"></a>

**Model type:** Logistic Regression\
**Dependent variable:** STATISTICAL_MURDER_FLAG\
**Explanatory variables:** BORO, PERP_AGE_GROUP, PERP_SEX, PERP_RACE,
VIC_AGE_GROUP, VIC_SEX, VIC_RACE

The primary objectives of the model are to:

-   **Identify the impact of factors on probability of a murder
    incident:**\
    Examine how various demographic factors influence the probability of
    a murder incident across different boroughs by examining the model
    coefficients.

-   **Assess factor significance:**\
    Identify which factors do not play a significant role in explaining
    the probability of a murder incident.

### 3.1 Summary <a class="anchor" id="chapter_3_1"></a>

```{r}
set.seed(0)
split_vector <- sample(c(TRUE, FALSE)
                       , size = nrow(data)
                       , replace = TRUE
                       , prob = c(0.8, 0.2))
data_train <- data[split_vector, ]
data_test <- data[!split_vector, ]
```

```{r}
model <- glm(STATISTICAL_MURDER_FLAG ~ BORO + PERP_AGE_GROUP + PERP_SEX + PERP_RACE + VIC_AGE_GROUP + VIC_SEX + VIC_RACE
             , data = data_train
             , family = binomial
             )
```

```{r}
summary(model)
```

### 3.2 Performance <a class="anchor" id="chapter_3_2"></a>

#### Model 10-fold cross validation mean accuracy: 80.6%

```{r}
get_mean_accuracy <- function(data, k){
    folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
    
    accuracy_values <- c()
    
    for (i in 1:k) {
        indices_test <- which(folds == i, arr.ind = TRUE)
        data_test <- data[indices_test, ]
        data_train <- data[-indices_test, ]
        data_test$VIC_AGE_GROUP <- factor(data_test$VIC_AGE_GROUP
                                          ,levels = levels(data_train$VIC_AGE_GROUP))
        data_test$PERP_AGE_GROUP <- factor(data_test$PERP_AGE_GROUP
                                          ,levels = levels(data_train$PERP_AGE_GROUP))
        model <- glm(STATISTICAL_MURDER_FLAG ~ BORO + PERP_AGE_GROUP + PERP_SEX + PERP_RACE + VIC_AGE_GROUP + VIC_SEX + VIC_RACE
                 , data = data_train
                 , family = binomial
                 )
        predicted_probabilities <- predict(model
                                           , newdata = data_test
                                           , type = "response")
        predicted_probabilities[is.na(predicted_probabilities)] <- 0.5
        predicted_classes <- predicted_probabilities > 0.5
        accuracy <- mean(predicted_classes == data_test$STATISTICAL_MURDER_FLAG)
        accuracy_values[i] <- accuracy
    }

    return(mean(accuracy_values))
}
```

```{r}
get_mean_accuracy(data, 10)
```

## 4. Bias <a class="anchor" id="chapter_4"></a>

### 4.1 Possible bias <a class="anchor" id="chapter_4_1"></a>

1.  **Incomplete Reporting:**\
    Underreporting or misclassification of incidents, which skews the
    overall data.

2.  **Model Specification:**\
    Omitting relevant variables or including irrelevant ones can lead to
    incorrect inferences about relationships within the data.

### 4.2 Dataset bias <a class="anchor" id="chapter_4_2"></a>

The dataset has around **43%** of unlabeled data regarding perpetrator
race.\
If this data is ommited, the dataset becomes highly imbalanced, with the
BLACK race prevailing at **73%** of the remaing data and the next,
second place, WHITE HISPANIC race, sitting at just **15%**.\
This skewed data distribution makes it challenging, if not impossible,
to conduct an unbiased analysis

```{r}
data_race <- data %>%
    group_by(PERP_RACE) %>%
    summarise(INCIDENT_COUNT = n(), .groups = 'drop') %>%
    mutate(PERCENTAGE = round(INCIDENT_COUNT/sum(INCIDENT_COUNT)*100,2)) %>%
    arrange(desc(INCIDENT_COUNT))
```

```{r}
data_race
```

```{r}
filter <- !(is.na(data_race$PERP_RACE) | data_race$PERP_RACE == '(null)' | data_race$PERP_RACE == 'UNKNOWN')
data_race_filtered <- data_race[filter,] %>%
    mutate(PERCENTAGE = round(INCIDENT_COUNT/sum(INCIDENT_COUNT)*100,2))
```

```{r}
data_race_filtered
```

## 5. Conclusion <a class="anchor" id="chapter_5"></a>

The analysis of the NYPD Shooting Incident Data (Historic) using
logistic regression reveals insights into the factors influencing the
likelihood of murder in shooting incidents across New York City. The
model incorporates various demographic and geographic variables,
including borough, age group, sex, and race of both perpetrators and
victims. After measuring model performance using 10-fold
cross-validation, the mean accuracy was found to be 80.6%.

Insights:

-   **Borough Influence:**\
    The analysis indicates that shootings in different boroughs have
    varying but statistically significant impacts on the likelihood of
    murder. Only Staten Island was found to be statistically
    insignificant as an explanatory variable, while the other boroughs
    were statistically significant (p-values \< 0.05).

-   **Perpetrator Age:**\
    The age group of the perpetrator plays a crucial role, particularly
    for those under 18 and those aged 18-24, both showing strong
    positive associations with murder likelihood (p-values \< 0.001).
    This suggests that younger individuals are more likely to be
    involved in fatal shooting incidents.

-   **Perpetrator Sex:**\
    The model shows that male perpetrators are significantly associated
    with higher odds of murder compared to female perpetrators. Both
    variables are statistically significant (p-values \< 0.001).

-   **Victim Demographics:**\
    Contrary to the demographic trends of perpetrators, where age, sex,
    and race are statistically significant, victim demographics indicate
    that only age plays a significant role in the probability of being
    murdered. Victims aged 25-64 show a strong association with murder
    incidents (p-values \< 0.001).

-   **Racial Dynamics:**\
    There is a notable imbalance in the racial distribution of
    perpetrators, with a high percentage of incidents involving Black
    individuals. However, due to approximately 43% of the data being
    unlabeled regarding perpetrator race, this could skew results and
    complicate unbiased analysis. The predominance of Black individuals
    in the remaining labeled data (73%) raises concerns about
    representation and bias in crime statistics.
